Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-22 papers/paper/ms.tex
update ms

{\it Near the main sequence (bit 9)}---Figure~\ref{fig:stellarprops} shows
that many stars observed by Kepler
are far from the main sequence.  Some of the challenges this introduces
for rotation-based ages include unresolved binaries, metallicity,
reddening, and drivers of rotation other than magnetic braking.
After exploring various options, we settled on the orange locus in the
$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}
as a way of flagging unresolved binaries, as well as evolved
late-F and early-G stars.  While the exact details of how this locus
is constructed are arbitrary (see Appendix~\ref{app:linemethod}), the
general aim is to flag stars for which there is evidence based on
their location in the \replaced{HR}{Kiel} diagram that our gyrochronology model may not
be reliable.  While the $\log g$--$T_{\rm eff}$ cut flags about half
of Kepler stars with rotation periods as potentially not being
suitable for gyrochronology, a few outliers in $M_{\rm G}$ vs.~$G_{\rm
BP}$-$G_{\rm RP}$ remained after applying it.  Such outliers are
likely also questionable.  We therefore also fitted a polynomial to
the KOI main sequence in $M_{\rm G}$ vs.~$G_{\rm
	BP}$-$G_{\rm RP}$, and flagged stars more than 1\,magnitude from
this locus as part of the same bitflag.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest
  systems, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode the $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  All quoted age uncertainties are statistical.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}^{\rm \ast}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{
  	EW$_{\rm Li}^{\rm \ast}$ is the lithium equivalent width {\it after} subtracting a constant 7.5\,m\AA\ to account for the \ion{Fe}{1} 6707.44\,\AA\ blend (see Section~\ref{subsec:lithiumsel}).
  	Two checkmarks ($\checkmark\checkmark$) denote confirmed planets with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are
  expected to be reliable.  One checkmark ($\checkmark$) denotes
  confirmed planets with two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before``candidate'' planets.  
  Planetary sizes are mostly drawn from (in order of precedence)
  \citet{Petigura_2022}, \citet{Berger_2020b_rpage}, and \citet{Thompson_2018}, and might be unphysical
  for grazing planets.
  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main sequence?
  Bit 10: \citetalias{Santos_2021} CP/CB?
  Bit 11: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest a $Q_{\rm star}$ bitmask with bits 0--9
  not raised, and for a planet to be ``reliable'', we suggest
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}

>>>>

{\it Near the main sequence (bit 9)}---Figure~\ref{fig:stellarprops} shows
that many stars observed by Kepler
are far from the main sequence.  Some of the challenges this introduces
for rotation-based ages include unresolved binaries, metallicity,
reddening, and drivers of rotation other than magnetic braking.
After exploring various options, we settled on the orange locus in the
$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}
as a way of flagging unresolved binaries, as well as evolved
late-F and early-G stars.  While the exact details of how this locus
is constructed are arbitrary (see Appendix~\ref{app:linemethod}), the
general aim is to flag stars for which there is evidence based on
their location in the \replaced{HR}{HR or Kiel} diagram\added{s} that our gyrochronology model may not
be reliable.  While the $\log g$--$T_{\rm eff}$ cut flags about half
of Kepler stars with rotation periods as potentially not being
suitable for gyrochronology, a few outliers in $M_{\rm G}$ vs.~$G_{\rm
BP}$-$G_{\rm RP}$ remained after applying it.  Such outliers are
likely also questionable.  We therefore also fitted a polynomial to
the KOI main sequence in $M_{\rm G}$ vs.~$G_{\rm
	BP}$-$G_{\rm RP}$, and flagged stars more than 1\,magnitude from
this locus as part of the same bitflag.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include the youngest
  systems, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode the $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  All quoted age uncertainties are statistical.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}^{\rm \ast}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{
  	EW$_{\rm Li}^{\rm \ast}$ is the lithium equivalent width {\it after} subtracting a constant 7.5\,m\AA\ to account for the \ion{Fe}{1} 6707.44\,\AA\ blend (see Section~\ref{subsec:lithiumsel}).
  	Two checkmarks ($\checkmark\checkmark$) denote confirmed planets with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are
  expected to be reliable.  One checkmark ($\checkmark$) denotes
  confirmed planets with two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before``candidate'' planets.  
  Planetary sizes are mostly drawn from (in order of precedence)
  \citet{Petigura_2022}, \citet{Berger_2020b_rpage}, and \citet{Thompson_2018}, and might be unphysical
  for grazing planets.
  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main sequence?
  Bit 10: Spectroscopic $|[{\rm Fe/H}]|$$>$0.3?
  Bit 11: \citetalias{Santos_2021} CP/CB?
  Bit 12: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 11 and bit 7.
  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest a $Q_{\rm star}$ bitmask with bits 0--10
  not raised, and for a planet to be ``reliable'', we suggest
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}
